---
title: Racial Competition and Charter School Expansion in California
author:
  - Richard Paquin Morel
  - Postdoctoral Research Associate
  - University of Pittsburgh
  - richard.morel@pitt.edu
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    reference_docx: bib/academic_template.docx
bibliography: bib/aefp_bib.bib
csl: bib/apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lfe)
library(huxtable)
library(broom)
```

While charter schools are often positioned as a means to reduce the opportunity gap between students of color and their white peers, they may  contribute to increasing racial segregation [@bifulcoSchoolChoiceRacial2007; @frankenbergNotJustUrban2012; @garciaImpactSchoolChoice2008; @rodaSchoolChoicePolicies2013] or expose students to different regimes of instruction or discipline [@golannParadoxSuccessNoexcuses2015]. Research also suggests that advantaged families often use choice to their own advantage [@lauenChooseNotChoose2009]. To what extent does this occur with charter schools? Existing reseach is limited, with conflicting findings [@renzulliSchoolChoiceCharter2005; @rappDispellingMythWhite2007], due to theoretical and methodological issues. 

To address limitations in the literature, I develop a conceptual framework, based on racial competition theory, to hypothesize that advantaged families may migrate to charter schools when exposed, proximally or directly, to members of non-dominant racial groups. The theory predicts that, under such conditions, members of white communities exhibit reactionary behaviors in order to protect control over social, political, and economic resources [@blalockTheoryMinorityGroup1967; @blumerRacePrejudiceSense1958; @boboPrejudiceGroupPosition1999; @oliverParadoxesIntegrationRace2010]. I test both a proximal threat hypothesis--increases in the share of non-whites in neighboring communities--and a direct threat hypothesis--increases in the share of non-whites within schools. In addition, I leverage a panel of data spanning nearly two decades and fixed-effects models to account for unobserved, non-time-varying differences between school districts.

## Data and Methods

I complied a panel of demographic, achievement, geographic, and finance data for all schools and districts in California from 2000-2016. These data were accessed through the Urban Institute’s Education Data Portal’s application programming interface and from the National Center for Education Statistics. I subset the panel to include only majority white districts, defined as those with a mean share of white students across all years in the fourth or fifth quantile for all districts. The resulting dataset has 8144 district-year observations and 448 unique school districts.

My main independent variables of interest are proxies for the proximal threat hypothesis and the direct threat hypothesis. To proxy proximal threat, I construct a “threat” variable [@andrewsGroupThreatPolicy2015; @cunninghamContextsMobilizationSpatial2007] that measures the extent to which members of dominant group face a proximal exposure to minority groups:

$$ Threat_{ij} = Threatened_{ij} * Threatening_{kj}$$
where _i_ indexes school districts, _j_ indexes years and _k_ indexes districts that share a border with district _i_. The status threat that parents in district _i_ are exposed to in year _j_ is a function of share of the threatened population _within_ in the district (here, white families) and the share of threatening populations in neighboring districts (here, black and Hispanic families).

The main specification I use is a district and year fixed-effects model:

$$ Y_{ij} = \beta_1Threat_{ij} + \beta_2Hisp_{trad} + \beta_3Blk_{trad} + X_{it}\beta + \alpha_i + \gamma_j + \epsilon_{ij} $$

where $Y_{ij}$ represents the (1) the share of charter schools in a district or (2) the share of white students enrolled in charter schools. I test a direct threat hypothesis by including the share of black and Hispanic students in traditional public schools and a proximal threat hypothesis by includeing the _threat_ variable. A vector of district-level time-varying controls are included (noted in the tables below).

## Preliminary Results

There is a strong association between the proximal and direct threat a district faces and its share of charters (Table 1) and white enrollment in charters (Table 2). As Table 3 shows, it is threat associated with increasing Hispanic populations in proximate districts that is driving the results in Tables 1 and 2. As a placebo test, I specified models with the share of Hispanic and Black students enrolled in charter schools. Threat should not impact whether or not these populations enroll in charter schools. Table 4 shows the results for these models, with threat showing no effect on enrollment for these groups.

## Significance

I find that threat, conceptualized both proximally and directly, contributes to both (1) charter schools open in predominately white areas and (2) white students enroll in charter schools. This finding extends our understanding of the role of charter schools in racial segregation, showing that families in predominately white areas may use charter schools to "escape" increasing diversity in traditional public schools. This is furtherevidence that charter schools may contribute to "white flight" from traditional public schools [@renzulliSchoolChoiceCharter2005]. As patterns of migration shift and suburban areas grow more diverse, advantaged families may seek out ways to exploit policies in order to maintain control over educational opportunities and resources.

```{r tab1}
load("ca_charter_analysis.RData")

m1 <-
  felm(
    prop_charter ~ 
      threat_overall +
      prop_hispanic_traditional +
      prop_black_traditional +
      prop_white_traditional  | 0 | 0 | leaid,
    data = ca_charter_maj_white
    )

m2 <-
  felm(
    prop_charter ~ 
      threat_overall + 
      prop_hispanic_traditional + 
      prop_black_traditional +
      prop_white_traditional +
      seg + 
      lag(prop_charter) +
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | 0 | 0 | leaid,
    data = ca_charter_maj_white
    )

m3 <-
  felm(
    prop_charter ~ 
      threat_overall + 
      prop_white_traditional +
      prop_hispanic_traditional + 
      prop_black_traditional +
      seg + 
      lag(prop_charter) +
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | leaid | 0 | leaid,
    data = ca_charter_maj_white
    )

m4 <-
  felm(
    prop_charter ~ 
      threat_overall + 
      prop_hispanic_traditional + 
      prop_white_traditional +
      prop_black_traditional +
      seg + 
      lag(prop_charter) +
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | leaid + year | 0 | leaid,
    data = ca_charter_maj_white
    )

models <- list(m1, m2, m3, m4)

tab <- 
  huxreg(
    models,
    coefs = 
      c(
        "Overall Threat" = "threat_overall",
        "%Hispanic students (TPS)"= "prop_hispanic_traditional",
        "%Black students (TPS)" = "prop_black_traditional"
      ),
    note = "{stars}. Includes controls for district segregation, the share of white students in the district, logged district enrollment, a lag for the share of district funding from local source and from property taxes, and lagged per pupil spending. Standard errors clustered at the district level. TPS = Traditional public schools."
    ) %>% 
  insert_row(c("Proximal threat", rep("", 4)), after = 1) %>% 
  insert_row(c("Direct threat", rep("", 4)), after = 4) %>% 
  insert_row(c("District Fixed Effect", "", "", "X", "X"), after = 9) %>% 
  insert_row(c("Year Fixed Effect", "", "", "", "X"), after = 10) %>%
  set_align(1:14, 1, "left") %>% 
  set_align(1:14, 2:5, "center") %>% 
  set_bottom_border(c(2, 10, 11), 1:5, 0) %>% 
  set_bold(c(2, 5), 1, TRUE) %>% 
  as_flextable() %>% 
  flextable::set_caption("Table 1. Estimates of the Effects of Proximate and Direct Threat on Share of Charter Schools in a District") %>% 
  flextable::padding(i = 1:14, j = 1:5, padding = 0)
  
tab
  
```


```{r tab2}
m1 <-
  felm(
    prop_white_charter ~ 
      threat_overall +
      prop_hispanic_traditional +
      prop_black_traditional +
      prop_white_traditional  | 0 | 0 | leaid,
    data = ca_charter_maj_white
    )

m2 <-
  felm(
    prop_white_charter ~ 
      threat_overall + 
      prop_hispanic_traditional + 
      prop_black_traditional +
      prop_white_traditional +
      seg + 
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | 0 | 0 | leaid,
    data = ca_charter_maj_white
    )

m3 <-
  felm(
    prop_white_charter ~ 
      threat_overall + 
      prop_white_traditional +
      prop_hispanic_traditional + 
      prop_black_traditional +
      seg + 
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | leaid | 0 | leaid,
    data = ca_charter_maj_white
    )

m4 <-
  felm(
    prop_white_charter ~ 
      threat_overall + 
      prop_hispanic_traditional + 
      prop_white_traditional +
      prop_black_traditional +
      seg + 
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | leaid + year | 0 | leaid,
    data = ca_charter_maj_white
    )

models <- list(m1, m2, m3, m4)

tab <- 
  huxreg(
    models,
    coefs = 
      c(
        "Overall Threat" = "threat_overall",
        "%Hispanic students (TPS)"= "prop_hispanic_traditional",
        "%Black students (TPS)" = "prop_black_traditional"
      ),
    note = "{stars}. Includes controls for district segregation, the share of white students in the district, logged district enrollment, a lag for the share of district funding from local source and from property taxes, and lagged per pupil spending. Standard errors clustered at the district level. TPS = traditional public schools."
    ) %>% 
  insert_row(c("Proximal threat", rep("", 4)), after = 1) %>% 
  insert_row(c("Direct threat", rep("", 4)), after = 4) %>% 
  insert_row(c("District Fixed Effect", "", "", "X", "X"), after = 9) %>% 
  insert_row(c("Year Fixed Effect", "", "", "", "X"), after = 10) %>%
  set_align(1:14, 1, "left") %>% 
  set_align(1:14, 2:5, "center") %>% 
  set_bottom_border(c(2, 10, 11), 1:5, 0) %>% 
  set_bold(c(2, 5), 1, TRUE) %>% 
  as_flextable() %>% 
  flextable::set_caption("Table 2. Estimates of the Effects of Proximate and Direct General Threat on White Charter School Enrollment") %>% 
  flextable::padding(i = 1:14, j = 1:5, padding = 0)
  
tab
  
```


```{r tab3}

m1 <-
  felm(
    prop_charter ~ 
      threat_hispanic + 
      threat_black +
      prop_hispanic_traditional + 
      prop_white_traditional +
      prop_black_traditional +
      seg + 
      lag(prop_charter) +
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | leaid + year | 0 | leaid,
    data = ca_charter_maj_white
    )

m2 <-
  felm(
    prop_white_charter ~ 
      threat_hispanic + 
      threat_black +
      prop_hispanic_traditional + 
      prop_white_traditional +
      prop_black_traditional +
      seg + 
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | leaid + year | 0 | leaid,
    data = ca_charter_maj_white
    )

tab <- 
  huxreg(
    "Share of charter schools" = m1, "White Enrollment in Charter schools" = m2,
    coefs = 
      c(
        "Hispanic threat" = "threat_hispanic", 
        "Black threat" = "threat_black",
        "%Hispanic students (TPS)"= "prop_hispanic_traditional",
        "%Black students (TPS)" = "prop_black_traditional"
      ),
    note = "{stars}. Includes controls for district segregation, the share of white students in the district, logged district enrollment, a lag for the share of district funding from local source and from property taxes, and lagged per pupil spending. Standard errors clustered at the district level. TPS = traditional public schools."
    ) %>% 
  insert_row(c("Proximal threat", rep("", 2)), after = 1) %>% 
  insert_row(c("Direct threat", rep("", 2)), after = 6) %>% 
  insert_row(c("District Fixed Effect", "X", "X"), after = 11) %>% 
  insert_row(c("Year Fixed Effect", "X", "X"), after = 12) %>% 
  set_bottom_border(c(2, 12, 13), 1:3, 0) %>% 
  set_align(2, 1, "left") %>% 
  set_align(1:16, 2:3, "center") %>% 
  set_bold(c(2, 7), 1, TRUE) %>% 
  as_flextable() %>% 
  flextable::set_caption("Table 3. Estimates of the Effects of Proximate and Direct Specific Threat on Charter Founding and Enrollment") %>% 
  flextable::padding(i = 1:16, j = 1:3, padding = 0)
  
tab
```


```{r tab4}
m1 <-
  felm(
    prop_black_charter ~ 
      threat_overall + 
      prop_hispanic_traditional + 
      prop_white_traditional +
      prop_black_traditional +
      seg + 
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | leaid + year | 0 | leaid,
    data = ca_charter_maj_white
    )

m2 <-
  felm(
    prop_hispanic_charter ~ 
      threat_overall + 
      prop_hispanic_traditional + 
      prop_white_traditional +
      prop_black_traditional +
      seg + 
      log(enrollment) +
      lag(prop_local) +
      lag(prop_local_prop_tax) + 
      lag(per_pupil_spending) | leaid + year | 0 | leaid,
    data = ca_charter_maj_white
    )

tab <- 
  huxreg(
    "Black Enrollment in Charter schools" = m1, "Hispanic Enrollment in Charter schools" = m2,
    coefs = 
      c(
        "Overall Threat" = "threat_overall",
        "%Hispanic students (TPS)"= "prop_hispanic_traditional",
        "%Black students (TPS)" = "prop_black_traditional"
      ),
    note = "{stars}. Includes controls for district segregation, the share of white students in the district, logged district enrollment, a lag for the share of district funding from local source and from property taxes, and lagged per pupil spending. Standard errors clustered at the district level. TPS = Traditional public schools."
    ) %>% 
  insert_row(c("Proximal threat", "", ""), after = 1) %>%
  insert_row(c("Direct threat", "", ""), after = 4) %>%
  insert_row(c("District Fixed Effect", "X", "X"), after = 9) %>% 
  insert_row(c("Year Fixed Effect", "X", "X"), after = 10) %>% 
  set_bottom_border(c(2, 10, 11), 1:3, 0) %>% 
  set_align(2, 1, "left") %>% 
  set_align(1:14, 2:3, "center") %>% 
  set_bold(c(2, 5), 1, TRUE) %>% 
  as_flextable() %>% 
  flextable::set_caption("Table 4. Estimates of the Effects of Proximate and Direct Threat on Black and Hispanic Charter School Enrollment") %>%
  flextable::padding(i = 1:14, j = 1:3, padding = 0)
  
tab
```


# References