---
title: Status Anxiety and Charter School Expansion in California
author:
  - name: Richard Paquin Morel
    affil: 1
affiliation:
  - num: 1
    address: Postdoctoral Research Associate, Learning Research and Development Center, University of Pittsburgh
column_numbers: 4
body_textsize: 35px
logoleft_name: RGB_Shield_Stacked_Education_3color.png 
output: 
  posterdown::posterdown_html:
    self_contained: false
    keep_md: true
bibliography: packages.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(tidycensus)
library(lfe)
library(sf)
library(patchwork)
library(huxtable)
library(dotwhisker)
library(cowplot)

load("../ca_threat_analysis.RData")

theme_set(
  theme_minimal() +
    theme(
      axis.title.x = element_text(size = 10, hjust = 1),
      axis.title.y = element_text(size = 10),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(size = 10, colour = "grey25", face = "bold"),
      plot.subtitle = element_text(size = 9, colour = "grey45"),
      legend.position = "bottom")
  )

charter_dat <- 
  charter_dat %>% 
  mutate_at(vars(matches("share|threat")), list(scaled = ~ . * 100)) %>% 
  mutate(prop_charter_scaled = prop_charter * 100,
         est_population_5_17_poverty_pct_scaled = est_population_5_17_poverty_pct * 100,
         isolation_scaled = isolation * 100)

charter_dat_non_urban <-
  charter_dat_non_urban %>% 
  mutate_at(vars(matches("share|threat")), list(scaled = ~ . * 100)) %>% 
  mutate(prop_charter_scaled = prop_charter * 100,
         est_population_5_17_poverty_pct_scaled = est_population_5_17_poverty_pct * 100,
         isolation_scaled = isolation * 100)

charter_dat_urban <-
  charter_dat_urban %>% 
  mutate_at(vars(matches("share|threat")), list(scaled = ~ . * 100)) %>% 
  mutate(prop_charter_scaled = prop_charter * 100,
         est_population_5_17_poverty_pct_scaled = est_population_5_17_poverty_pct * 100,
         isolation_scaled = isolation * 100)
```

# Overview

While school choice policies are now ubiqitous are often presented as a strategy for addressing persistent racial disparities education, there is concern that such policies may maintain or increases such disparities due to resource hoarding by advantaged families or through increased racial segregation [@davisSchoolChoiceSegregation2014; rodaSchoolChoicePolicies2013]. More specifically, white families may use charters to exit from racially diversifying schools to maintain status hierarchies [@renzulliSchoolChoiceCharter2005]. Using a group threat/status anxiety framework [@boboPerceptionsRacialGroup1996], I hypothesize that exposure to increased racial diversification may induce white families to enroll their children in charter schools. I distinguish between proximal exposure, where racial diversification occurs in neighboring school districts, and direct exposure, where racial diversification occurs within local traditional public schools. 

I find evidence that: 

1. both proximal and direct exposure are associated with both **increased foundings of charter schools** and with **white enrollment in charter schools**;  
2. this is concentrated in **non-urban areas**; and  
3. threat-induced white enrollment is associated with **increased racial segregation**.  

# Data
Panel of California school district data from 2000-2015 (n = 16693 district-by-year observations; 1111 unique districts)

Sources: The Common Core of Data; Small Area Income and Poverty Estimates (both accessed via the Urban Institutes API); NCES EDGE database

Charter schools in California:

- Law enacted in 1992
- Authorization occurs at the district level
- In 2000, ~ 2% of students were enrolled in charters
- By 2015, ~ 10% were enrolled in charters

# Methods

**Measures**

1. _Proximal status threat_ [@andrewsGroupThreatPolicy2015]
$$ProxThreat_{ij} = \theta_{ij} * \gamma_{ij}$$
where $\theta_{ij} = \frac{White_{ij}}{Total_{ij}}$ and $\gamma_{ij} = \frac{\sum_1^k Blk_{kj}, Hisp_{kj}}{\sum_1^k Total_{kj}}$

2. _Direct status threat_
$$DirThreat_{ij} = \frac{\{BlkTPS{ij}, HispTPS{ij}\}}{\{TotalBlk_{ij},TotalHisp_{ij}\}}$$


3. _Racial segregation (white isolation)_ [@masseySuburbanizationSegregationMetropolitan1988]
$$WhiteIsolation_{ij} = \sum_{s=1}^n[(\frac{x_{sj}}{X_{ij}})(\frac{x_{sj}}{t_{ij}})]$$
where $x_{sj}$ is the white enrollment in a school, $X_{ij}$ is the total white enrollment in the district $s$ is in, and $t_{ij}$ is the total enrollment in the district.

**Analytic models**

1. Does proximal and/or direct exposure increase charter enrollment for white students?  
$$Y_{ij} = \beta_1ProxThreat_{ij} + \beta_2DirThreat_{ij} + X_{ij}\beta + \alpha_i + \gamma_j + \epsilon_{ij}$$

2. Does white enrollment induced by exposure contribute to racial segregation in schools?  

_First stage_  
$$\phi_{ij} = \pi_1ProxThreat_{ij} + \pi_1DirThreat_{ij}$$

_Second stage_: 
$$WhiteIsolation_{ij} = \beta_1\hat{\phi} + X_{ij}\beta + \alpha_i + \gamma_j + \epsilon_{ij}$$


# Charter School Expansion and Enrollment

```{r map1, out.width='95%', fig.align='center', fig.height=5, dpi=340}
bivariate_color_scale <- tibble(
  group = c(
    "3 - 3", # high charter enrollment, high threat
    "2 - 3", # medium charter enrollment, high threat
    "1 - 3", # low charter enrollment, high threat
    "3 - 2", # high charter enrollment, med threat
    "2 - 2", # medium charter enrollment, med threat
    "1 - 2", # low charter enrollment, med threat
    "3 - 1", # high charter enrollment, low threat
    "2 - 1", # medium charter enrollment, low threat
    "1 - 1"  # low charter enrollment, low threat
  ),
  color = rev(pals::stevens.pinkblue())
)

ca_ds <- list(shapefiles[[1]] %>% mutate(year = year + 1), shapefiles[[9]])

ca_threat_seg <- charter_dat %>%  
  filter(!is.na(threat_overall), 
         !is.na(white_share_within_charter),
         year %in% c(2000, 2015)) %>% 
  arrange(leaid, year) %>%  
  group_by(leaid) %>% 
  mutate(increase_threat = as.numeric(threat_overall - lag(threat_overall) > 0), 
         increase_white = as.numeric(white_share_within_charter- lag(white_share_within_charter) > 0)) %>% 
  fill(increase_threat, increase_white, .direction = "up") %>%
  mutate(increase_both = increase_threat*increase_white) %>% 
  ungroup() %>% 
  group_split(year) %>% 
  map2(
    ca_ds, 
    ~ .x %>% 
      mutate(
        white_enroll = ntile(white_share_within_charter, 3),
        threat_quantiles = ntile(threat_overall, 3)
      ) %>% 
      select(leaid, year,  increase_threat:last_col()) %>% 
      mutate(group = paste(as.numeric(white_enroll), "-", as.numeric(threat_quantiles))) %>% 
      left_join(bivariate_color_scale) %>% 
      left_join(.y %>% 
                  select(geoid, geometry),
                by = c("leaid" = "geoid"))
  )

ca_threat_seg <- rbind(ca_threat_seg[[1]], ca_threat_seg[[2]])

ca_map <- ca_threat_seg %>% 
  ggplot() +
  geom_sf(aes(geometry = geometry, fill = color), color = NA) +
  scale_fill_identity() +
  geom_sf(data = ca_threat_seg %>% filter(increase_both == 1), aes(geometry = geometry, color = factor(increase_both)), size = 0.3, fill = "transparent", inherit.aes = FALSE) + 
  scale_color_manual(name = "", values = "magenta", labels = "Increased threat and white enrollment") +
  labs(title = "Proximal Threat and White Charter School Enrollment in California") +
  theme(legend.position = "bottom", 
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  coord_sf() +
  facet_wrap(~ year)


leg <-
  bivariate_color_scale  %>% 
  separate(group, into = c("rev", "white"), sep = " - ") %>%
  mutate(rev = as.integer(rev),
         white = as.integer(white)) %>% 
  ggplot() +
  geom_tile(
    mapping = aes(
      x = rev,
      y = white,
      fill = color)
  ) +
  scale_fill_identity() +
  labs(x = "Greater white charter enrollment \u2192",
       y = "Greater proximal threat \u2192") + 
  theme_void() +
  # make font small enough
  theme(
    axis.title = element_text(size = 5),
    axis.title.y = element_text(angle = 90, size = 5),
    axis.title.x = element_text(size = 5, hjust = 0)
  ) +
  # quadratic tiles
  coord_fixed()

p1 <- charter_dat %>% 
  filter(year %in% c(2000, 2015)) %>% 
  right_join(rbind(shapefiles[[1]] %>% mutate(year = year + 1), shapefiles[[9]]), by = c("leaid" = "geoid", "year")) %>% 
  ggplot() + 
  geom_sf(aes(geometry = geometry, fill = prop_charter_scaled), color = NA) + 
  facet_wrap(~ year, nrow = 1) +
  scale_fill_viridis_c(name = "", na.value = "white", begin = 0.1, end = 1) + 
  labs(title = "District Share of Charter Schools") +
  theme(legend.position = "bottom", 
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

p2 <- charter_dat %>% 
  filter(year %in% c(2000, 2015)) %>% 
  right_join(rbind(shapefiles[[1]] %>% mutate(year = year + 1), shapefiles[[9]]), by = c("leaid" = "geoid", "year")) %>% 
  ggplot() + 
  geom_sf(aes(geometry = geometry, fill = white_share_within_charter_scaled), color = NA) + 
  facet_wrap(~ year, nrow = 1) + 
  scale_fill_viridis_c(name = "", na.value = "white", begin = 0.1, end = 1) + 
  labs(title = "Share of White Students Enrolled in Charter Schools") +
  theme(legend.position = "bottom", 
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

p1 
p2
ggdraw(ca_map) +
  draw_plot(leg, 0, 0, 0.25, 0.25, scale = 0.7)
```

# Descriptive Analysis

```{r out.width='90%', fig.align='center', dpi=340}
#charter_dat %>% 
#  filter(n_charter > 0) %>% 
#  mutate(urbanicity = ifelse(urban_centric_locale %in% c(1, 2, 11, 12), "Urban Districts", "Non-Urban Districts")) %>% 
#  ggplot(aes(x = car::logit(threat_overall), y = car::logit(white_share_within_charter))) + 
#  geom_point(aes(color = urbanicity, size = enrollment), alpha = 0.5) + 
#  geom_smooth(aes(color = urbanicity), method = "lm", size = 0.5, se = FALSE) + 
#  labs(x = "Share of white students enrolled in charters (logit transformed)", 
#       y = "Proximal threat/exposure (logit transformed)",
#       title = "Relationship between white enrollment in charters and proximal threat by year",
#       caption = "Note. Includes districts with at least one charter school. Fitted lines based on linear model.") +
#  scale_color_viridis_d(name = "Urbanicity", begin = 0.25, end = 0.75) + 
#  scale_size_continuous(name = "Overall district enrollment", range = c(1, 4), labels = scales::comma) + 
#  facet_wrap(~ year)
#
#charter_dat %>% 
#  filter(n_charter > 0) %>% 
#  mutate(urbanicity = ifelse(urban_centric_locale %in% c(1, 2, 11, 12), "Urban Districts", "Non-Urban Districts")) %>% 
#  ggplot(aes(x = car::logit(threat_overall), y = car::logit(isolation))) + 
#  geom_point(aes(color = urbanicity, size = white_overall_district_share), alpha = 0.5) + 
#  geom_smooth(aes(color = urbanicity), method = "lm", size = 0.5, se = FALSE) + 
#  labs(x = "Proximal threat/exposure (logit transformed)",
#       y = "Share of white students enrolled in charters (logit trasformed)", 
#       title = "Relationship between proximal threat and white isolation by year",
#       caption = "Note. Includes districts with at least one charter school. Fitted lines based on linear model.") +
#  scale_color_viridis_d(name = "Urbanicity", begin = 0.25, end = 0.75) + 
#  scale_size_continuous(name = "White district enrollment", range = c(1, 4), labels = scales::comma) + 
#  facet_wrap(~ year)

charter_dat %>% 
  ggplot(aes(x = threat_overall_scaled)) +
  geom_histogram(bins = 50, fill = "cornflowerblue", color = "midnightblue", alpha = 0.7) +
  labs(x = "", y = "", title = "Distribution of proximal threat by year") +
  facet_wrap(~ year)

charter_dat %>% 
  select(year, black_share_btw_traditional, hispanic_share_btw_traditional) %>% 
  pivot_longer(cols = -year) %>% 
  mutate(name = fct_rev(as_factor(name))) %>% 
  ggplot(aes(x = value, group = name)) +
  geom_histogram(aes(x = value, fill = name, alpha = name, stat(ndensity)), bins = 50, color = "midnightblue") +
  scale_fill_manual(name = "", values = c("cornflowerblue", "magenta"), labels = c("%TPS students - Latinx", "%TPS students - Black")) +
  scale_alpha_manual(guide = FALSE, values = c(0.4, 0.9)) +
  labs(x = "", y = "", title = "Distribution of direct exposure by year") +
  facet_wrap(~ year)
```


# Preliminary Results

```{r standard-plot, out.width='100%', fig.align='center', dpi=340}
m1_controls <- 
  felm(
    prop_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat
  )

m2_controls <- 
  felm(
    white_share_within_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled| leaid + year | 0 | leaid,
    data = charter_dat
  )

models <- list("Share of Charters" = m1_controls, "White Enrollment in Charters" = m2_controls) 
mod_names <- names(models)
tidy_models <- 
  map(models, tidy) %>% 
  map(~ filter(.x, str_detect(term, "threat_|black|hispanic"))) %>% 
  map2(mod_names, ~ mutate(.x, model = .y)) %>% 
  map(~ mutate(.x, term = case_when(
    term == "threat_overall_scaled" ~ "Proximal Exposure",
    term == "black_share_btw_traditional_scaled" ~ "Direct (%Black - TPS)",
    term == "hispanic_share_btw_traditional_scaled" ~ "Direct (%Hispanic - TPS)"
  )))

tidy_models %>% 
  bind_rows() %>% 
  dwplot() +
  geom_vline(
    xintercept = 0, 
    colour = "grey60", 
    linetype = 2) + 
  scale_color_brewer(name = "", palette = "Paired", direction = -1) + 
  labs(y = "", x = "Coefficient Estimate",
       title = "Results of Direct and Proximal Threat on Charter Density and White Enrollment", 
       caption = str_wrap("TPS = Traditional public schools. Coefficient estimates based on OLS regression with district and year fixed effects. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of children in poverty; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, white students, and magnet schools in the district; and teacher-student ratio.", 120)) +
  theme(legend.position = "bottom") +
  coord_flip()

tab <- huxreg(models, 
       coefs = c("Proximal threat" = "threat_overall_scaled",
                 "%Black(TPS)" = "black_share_btw_traditional_scaled",
                 "%Hispanic(TPS)" = "hispanic_share_btw_traditional_scaled"),
       statistics = c('N' = 'nobs', 'R-squared' = 'r.squared'),
       note = "{stars}. TPS = Traditional public schools. Coefficient estimates based on OLS regression with district and year fixed effects. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of children in poverty; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, white students, and magnet schools in the district; and teacher-student ratio.") %>% 
  insert_row(c("District Fixed Effects", "X", "X"), after = 7) %>% 
  insert_row(c("Year Fixed Effects", "X", "X"), after = 8) %>% 
  insert_row(c("Controls", "X", "X"), after = 9) %>% 
  set_bottom_border(8, 2:3, 0) %>% 
  set_bottom_border(9, 2:3, 0) %>% 
  set_bottom_border(10, 2:3, 0) %>% 
  set_align(1:13, 2:3, "center") %>% 
  insert_row(c("Table 1: Fixed effects regression of share of charters and white enrollment \non proximal and direct threat", "", ""), after = 0) %>% 
  set_colspan(1, 1, 3) %>% 
  set_top_border(1, 1:3, 0) %>% 
  set_align(1, 1, "left") %>% 
  set_italic(1, 1, TRUE) %>% 
  set_wrap(1, 1, TRUE)
```

```{r morefigs, out.width='100%', fig.align='center', dpi=340}
m1_controls <- 
  felm(
    white_share_within_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat_non_urban
  )

m2_controls <- 
  felm(
    white_share_within_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat_urban
  )

m3_controls <- 
  felm(
    prop_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat_non_urban
  )

m4_controls <- 
  felm(
    prop_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat_urban
  )


models <- list("Non-urban white enrollment" = m1_controls, "Urban white enrollment" = m2_controls, 
               "Non-urban %Charter" = m3_controls, "Urban %Charter" = m4_controls)
mod_names <- names(models)
tidy_models <- 
  map(models, tidy) %>% 
  map(~ filter(.x, str_detect(term, "threat_|black|hispanic"))) %>% 
  map2(mod_names, ~ mutate(.x, model = .y)) %>% 
  map(~ mutate(.x, term = case_when(
    term == "threat_overall_scaled" ~ "Proximal Exposure",
    term == "black_share_btw_traditional_scaled" ~ "Direct (%Black - TPS)",
    term == "hispanic_share_btw_traditional_scaled" ~ "Direct (%Hispanic - TPS)"
  )))

tidy_models %>% 
  bind_rows() %>% 
  dwplot() +
  geom_vline(
    xintercept = 0, 
    colour = "grey60", 
    linetype = 2) + 
  scale_color_brewer(name = "", palette = "Paired", direction = -1) + 
  labs(y = "", x = "Coefficient Estimate",
       title = "Comparing white charter enrollment in non-urban and urban districts",
       caption = str_wrap("TPS = Traditional public schools. Coefficient estimates based on OLS regression with district and year fixed effects. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of children in poverty; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, white students, and magnet schools in the district; and teacher-student ratio.", 120)) +
  theme(legend.position = "bottom") +
  coord_flip()

tab <- huxreg(models, 
       coefs = c("Proximal threat" = "threat_overall_scaled", 
                 "%Black(TPS)" = "black_share_btw_traditional_scaled",
                 "%Hispanic(TPS)" = "hispanic_share_btw_traditional_scaled"),
       statistics = c('N' = 'nobs', 'R-squared' = 'r.squared'),
       note = "{stars}. TPS = Traditional public schools. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of 5-17 in poverty; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, and magnet schools in the district; and teacher-student ratio.") %>% 
  insert_row(c("District Fixed Effects", "X", "X", "X", "X"), after = 7) %>% 
  insert_row(c("Year Fixed Effects", "X", "X", "X", "X"), after = 8) %>% 
  insert_row(c("Controls", "X", "X", "X", "X"), after = 9) %>% 
  set_bottom_border(8, 2:5, 0) %>% 
  set_bottom_border(9, 2:5, 0) %>% 
  set_bottom_border(10, 2:5, 0) %>% 
  insert_row(c("", "Share of Charters", "", "White Enrollment in Charters", ""), after = 0) %>% 
  set_colspan(1, c(2, 4), 2) %>% 
  set_bottom_border(1, 1:5, 1) %>% 
  set_bottom_border(1, 1:5, 0) %>% 
  set_top_border(2, 1:5, 0) %>%
  set_align(1:14, 2:5, "center") %>% 
  insert_row(c("Table 2: Results for urban and non-urban school districts", 
               "", "", "", ""), after = 0) %>% 
  set_colspan(1, 1, 5) %>% 
  set_top_border(1, 1:5, 0) %>% 
  set_align(1, 1, "left") %>% 
  set_italic(1, 1, TRUE) %>% 
  set_wrap(1, 1, TRUE)
```

```{r out.width='100%', fig.align='center', dpi=340}
m1 <- 
  felm(
    isolation_scaled ~
      black_overall_district_share + 
      hispanic_overall_district_share + 
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct | leaid + year |  (white_share_within_charter_scaled ~ threat_overall_scaled) | leaid,
    data = charter_dat
  )

m2 <- 
  felm(
    isolation_scaled ~
      black_overall_district_share + 
      hispanic_overall_district_share + 
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct | leaid + year |  (white_share_within_charter_scaled ~ hispanic_share_btw_traditional_scaled + black_share_btw_traditional_scaled) | leaid,
    data = charter_dat
  )

m3 <- 
  felm(
    isolation_scaled ~
      black_overall_district_share + 
      hispanic_overall_district_share + 
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct | leaid + year |  (white_share_within_charter_scaled ~ threat_overall_scaled + black_share_btw_traditional_scaled + hispanic_share_btw_traditional_scaled) | leaid,
    data = charter_dat
  )

models <- list("Proximal Exposure" = m1, "Direct Exposure" = m2, 
               "Proximal and Direct Exposure" = m3)
mod_names <- names(models)
tidy_models <- 
  map(models, tidy) %>% 
  map(~ filter(.x, str_detect(term, "fit"))) %>% 
  map2(mod_names, ~ mutate(.x, model = .y)) %>% 
  map(~ mutate(.x, term = "Fitted Share of White Students in Charters"
  ))

tidy_models %>% 
  bind_rows() %>% 
  dwplot() +
  geom_vline(
    xintercept = 0, 
    colour = "grey60", 
    linetype = 2) + 
  scale_color_brewer(name = "", palette = "Paired", direction = -1) + 
  labs(y = "", x = "Coefficient Estimate",
       title = "Racial Isolation and Threat-induced White Enrollment in Charters",
       caption = str_wrap("Coefficient estimates based on IV regression with district and year fixed effects. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, and magnet schools in the district; and teacher-student ratio.", 120)) +
  theme(legend.position = "bottom") +
  coord_flip()
```


# Summary and implications

- Both direct and proximal exposure to students of color appear to motivate white students to leave traditional public schools for charter schools
- This process appears to increase district segregation--white students are more isolated from students of color
- Findings suggest that white families may use charter schools to maintain exclusive access to educational goods--aligning with other research showing that white families have a preference for schools with more white students

# References
