---
title: Status Anxiety and Charter School Expansion in California
author:
  - name: Richard Paquin Morel
    affil: 1
affiliation:
  - num: 1
    address: Postdoctoral Research Associate, Learning Research and Development Center, University of Pittsburgh
column_numbers: 4
font_family: Minion Pro
body_textsize: 35px
logoleft_name: RGB_Shield_Stacked_Education_3color.png 
output: 
  posterdown::posterdown_html:
    self_contained: true
    keep_md: true
bibliography: packages.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(tidycensus)
library(lfe)
library(sf)
library(patchwork)
library(huxtable)
library(dotwhisker)
library(cowplot)

load("../ca_threat_analysis.RData")

theme_set(
  theme_minimal() +
    theme(
      axis.title.x = element_text(size = 10, hjust = 1),
      axis.title.y = element_text(size = 10),
      axis.text.x = element_text(size = 8),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(size = 10, colour = "grey25", face = "bold"),
      plot.subtitle = element_text(size = 9, colour = "grey45"))
  )

charter_dat <- 
  charter_dat %>% 
  mutate_at(vars(matches("share|threat")), list(scaled = ~ . * 100)) %>% 
  mutate(prop_charter_scaled = prop_charter * 100,
         est_population_5_17_poverty_pct_scaled = est_population_5_17_poverty_pct * 100,
         isolation_scaled = isolation * 100)

charter_dat_non_urban <-
  charter_dat_non_urban %>% 
  mutate_at(vars(matches("share|threat")), list(scaled = ~ . * 100)) %>% 
  mutate(prop_charter_scaled = prop_charter * 100,
         est_population_5_17_poverty_pct_scaled = est_population_5_17_poverty_pct * 100,
         isolation_scaled = isolation * 100)

charter_dat_urban <-
  charter_dat_urban %>% 
  mutate_at(vars(matches("share|threat")), list(scaled = ~ . * 100)) %>% 
  mutate(prop_charter_scaled = prop_charter * 100,
         est_population_5_17_poverty_pct_scaled = est_population_5_17_poverty_pct * 100,
         isolation_scaled = isolation * 100)
```

# Overview

While school choice policies are now ubiqitous are often presented as a strategy for addressing persistent racial disparities education, there is concern that such policies may maintain or increases such disparities due to resource hoarding by advantaged families or through increased racial segregation. More specifically, white families may use charters to exit from racially diversifying schools. Using a group threat/status anxiety framework, I hypothesize that increased racial diversification may induce white families to enroll their children in charter schools. I distinguish between proximal status threat, where racial diversification occurs in neighboring school districts, and direct status threat, where racial diversification occurs within local traditional public schools. I find evidence that both proximal and direct threat are associated with both increased foundings of charter schools and with white enrollment in charter schools. 

# Data & Methods

**Data**. Panel of California school district data from 2000-2015 from the Common Core of Data, Small Area Income and Poverty Estimates (both accessed via the Urban Institutes API), and district boundary data from the NCES EDGE database

**Proximal status threat**
$$ProxThreat_{ij} = \theta_{ij} * \gamma_{ij}$$
where $\theta_{ij} = \frac{White_{ij}}{Total_{ij}}$ and $\gamma_{ij} = \frac{\sum_1^k Blk_{kj}, Hisp_{kj}}{\sum_1^k Total_{kj}}$

**Direct status threat**
$$DirThreat_{ij} = \frac{\{BlkTPS{ij}, HispTPS{ij}\}}{\{TotalBlk_{ij},TotalHisp_{ij}\}}$$

**White isolation**: 
$$WhiteIsolation_{ij} = \sum_{s=1}^n[(\frac{x_{sj}}{X_{ij}})(\frac{x_{sj}}{t_{ij}})]$$
where $x_{sj}$ is the white enrollment in a school, $X_{ij}$ is the total white enrollment in the district $s$ is in, and $t_{ij}$ is the total enrollment in the district.

- For each, $i$ indexes districts, $j$ indexes years, and $k$ indexes districts that neighbor $i$, and $s$ indexes schools


**Analytic models**

Does proximal and/or direct exposure increase charter enrollment for white students?  
$$Y_{ij} = \beta_1ProxThreat_{ij} + \beta_2DirThreat_{ij} + X_{ij}\beta + \alpha_i + \gamma_j + \epsilon_{ij}$$

Does white enrollment induced by exposure contribute to racial segregation in schools?  

_First stage_  
$$\phi_{ij} = \pi_1ProxThreat_{ij} + \pi_1DirThreat_{ij}$$

_Second stage_: 
$$WhiteIsolation_{ij} = \beta_1\hat{\phi} + X_{ij}\beta + \alpha_i + \gamma_j + \epsilon_{ij}$$


# Preliminary Results

```{r map1, out.width='100%', fig.align='left', fig.height=5, dpi=300}
bivariate_color_scale <- tibble(
  group = c(
    "3 - 3", # high charter enrollment, high threat
    "2 - 3", # medium charter enrollment, high threat
    "1 - 3", # low charter enrollment, high threat
    "3 - 2", # high charter enrollment, med threat
    "2 - 2", # medium charter enrollment, med threat
    "1 - 2", # low charter enrollment, med threat
    "3 - 1", # high charter enrollment, low threat
    "2 - 1", # medium charter enrollment, low threat
    "1 - 1"  # low charter enrollment, low threat
  ),
  color = rev(pals::stevens.pinkblue())
)

ca_ds <- list(shapefiles[[1]] %>% mutate(year = year + 1), shapefiles[[9]])

ca_threat_seg <- charter_dat %>%  
  filter(!is.na(threat_overall), 
         !is.na(white_share_within_charter),
         year %in% c(2000, 2015)) %>% 
  arrange(leaid, year) %>%  
  group_by(leaid) %>% 
  mutate(increase_threat = as.numeric(threat_overall - lag(threat_overall) > 0), 
         increase_white = as.numeric(white_share_within_charter- lag(white_share_within_charter) > 0)) %>% 
  fill(increase_threat, increase_white, .direction = "up") %>%
  mutate(increase_both = increase_threat*increase_white) %>% 
  ungroup() %>% 
  group_split(year) %>% 
  map2(
    ca_ds, 
    ~ .x %>% 
      mutate(
        white_enroll = ntile(white_share_within_charter, 3),
        threat_quantiles = ntile(threat_overall, 3)
      ) %>% 
      select(leaid, year,  increase_threat:last_col()) %>% 
      mutate(group = paste(as.numeric(white_enroll), "-", as.numeric(threat_quantiles))) %>% 
      left_join(bivariate_color_scale) %>% 
      left_join(.y %>% 
                  select(geoid, geometry),
                by = c("leaid" = "geoid"))
  )

ca_threat_seg <- rbind(ca_threat_seg[[1]], ca_threat_seg[[2]])

ca_map <- ca_threat_seg %>% 
  ggplot() +
  geom_sf(aes(geometry = geometry, fill = color), size = 0.1, color = "white") +
  scale_fill_identity() +
  geom_sf(data = ca_threat_seg %>% filter(increase_both == 1), aes(geometry = geometry, color = factor(increase_both)), size = 0.3, fill = "transparent", inherit.aes = FALSE) + 
  scale_color_manual(name = "", values = "magenta", labels = "Increased threat and white enrollment") +
  labs(title = "Proximal Threat and White Charter School Enrollment in California") +
  theme(legend.position = "bottom", 
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  coord_sf() +
  facet_wrap(~ year)


leg <-
  bivariate_color_scale  %>% 
  separate(group, into = c("rev", "white"), sep = " - ") %>%
  mutate(rev = as.integer(rev),
         white = as.integer(white)) %>% 
  ggplot() +
  geom_tile(
    mapping = aes(
      x = rev,
      y = white,
      fill = color)
  ) +
  scale_fill_identity() +
  labs(x = "Greater white charter enrollment \u2192",
       y = "Greater proximal threat \u2192") + 
  theme_void() +
  # make font small enough
  theme(
    axis.title = element_text(size = 5),
    axis.title.y = element_text(angle = 90, size = 5),
    axis.title.x = element_text(size = 5, hjust = 0)
  ) +
  # quadratic tiles
  coord_fixed()

ggdraw(ca_map) +
  draw_plot(leg, 0, 0, 0.25, 0.25, scale = 0.7)

p1 <- charter_dat %>% 
  filter(year %in% c(2000, 2015)) %>% 
  right_join(rbind(shapefiles[[1]] %>% mutate(year = year + 1), shapefiles[[9]]), by = c("leaid" = "geoid", "year")) %>% 
  ggplot() + 
  geom_sf(aes(geometry = geometry, fill = prop_charter_scaled), color = NA) + 
  facet_wrap(~ year, nrow = 1) +
  scale_fill_viridis_c(name = "", na.value = "white", begin = 0.1, end = 1) + 
  labs(title = "District Share of Charter Schools") +
  theme(legend.position = "bottom")

p2 <- charter_dat %>% 
  filter(year %in% c(2000, 2015)) %>% 
  right_join(rbind(shapefiles[[1]] %>% mutate(year = year + 1), shapefiles[[9]]), by = c("leaid" = "geoid", "year")) %>% 
  ggplot() + 
  geom_sf(aes(geometry = geometry, fill = white_share_within_charter_scaled), color = NA) + 
  facet_wrap(~ year, nrow = 1) + 
  scale_fill_viridis_c(name = "", na.value = "white", begin = 0.1, end = 1) + 
  labs(title = "Share of White Students Enrolled in Charter Schools") +
  theme(legend.position = "bottom")
```

```{r standard-plot, out.width='90%', fig.align='center', dpi=340}
m1_controls <- 
  felm(
    prop_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat
  )

m2_controls <- 
  felm(
    white_share_within_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled| leaid + year | 0 | leaid,
    data = charter_dat
  )

models <- list("Share of Charters" = m1_controls, "White Enrollment in Charters" = m2_controls) 
mod_names <- names(models)
tidy_models <- 
  map(models, tidy) %>% 
  map(~ filter(.x, str_detect(term, "threat_|black|hispanic"))) %>% 
  map2(mod_names, ~ mutate(.x, model = .y)) %>% 
  map(~ mutate(.x, term = case_when(
    term == "threat_overall_scaled" ~ "Proximal Exposure",
    term == "black_share_btw_traditional_scaled" ~ "Direct (%Black - TPS)",
    term == "hispanic_share_btw_traditional_scaled" ~ "Direct (%Hispanic - TPS)"
  )))

tidy_models %>% 
  bind_rows() %>% 
  dwplot() +
  geom_vline(
    xintercept = 0, 
    colour = "grey60", 
    linetype = 2) + 
  scale_color_brewer(name = "", palette = "Paired", direction = -1) + 
  labs(y = "", x = "Coefficient Estimate",
       title = "Results of Direct and Proximal Threat on Charter Density and White Enrollment", 
       caption = str_wrap("TPS = Traditional public schools. Coefficient estimates based on OLS regression with district and year fixed effects. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of children in poverty; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, white students, and magnet schools in the district; and teacher-student ratio.", 80)) +
  coord_flip()

tab <- huxreg(models, 
       coefs = c("Proximal threat" = "threat_overall_scaled",
                 "%Black(TPS)" = "black_share_btw_traditional_scaled",
                 "%Hispanic(TPS)" = "hispanic_share_btw_traditional_scaled"),
       statistics = c('N' = 'nobs', 'R-squared' = 'r.squared'),
       note = "{stars}. TPS = Traditional public schools. Coefficient estimates based on OLS regression with district and year fixed effects. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of children in poverty; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, white students, and magnet schools in the district; and teacher-student ratio.") %>% 
  insert_row(c("District Fixed Effects", "X", "X"), after = 7) %>% 
  insert_row(c("Year Fixed Effects", "X", "X"), after = 8) %>% 
  insert_row(c("Controls", "X", "X"), after = 9) %>% 
  set_bottom_border(8, 2:3, 0) %>% 
  set_bottom_border(9, 2:3, 0) %>% 
  set_bottom_border(10, 2:3, 0) %>% 
  set_align(1:13, 2:3, "center") %>% 
  insert_row(c("Table 1: Fixed effects regression of share of charters and white enrollment \non proximal and direct threat", "", ""), after = 0) %>% 
  set_colspan(1, 1, 3) %>% 
  set_top_border(1, 1:3, 0) %>% 
  set_align(1, 1, "left") %>% 
  set_italic(1, 1, TRUE) %>% 
  set_wrap(1, 1, TRUE)
```

```{r morefigs, out.width='90%', fig.align='center', dpi=340}
m1_controls <- 
  felm(
    white_share_within_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat_non_urban
  )

m2_controls <- 
  felm(
    white_share_within_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat_urban
  )

m3_controls <- 
  felm(
    prop_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat_non_urban
  )

m4_controls <- 
  felm(
    prop_charter_scaled ~
      threat_overall_scaled +
      black_share_btw_traditional_scaled +
      hispanic_share_btw_traditional_scaled +
      white_overall_district_share_scaled +
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct_scaled | leaid + year | 0 | leaid,
    data = charter_dat_urban
  )


models <- list("Non-urban white enrollment" = m1_controls, "Urban white enrollment" = m2_controls, 
               "Non-urban %Charter" = m3_controls, "Urban %Charter" = m4_controls)
mod_names <- names(models)
tidy_models <- 
  map(models, tidy) %>% 
  map(~ filter(.x, str_detect(term, "threat_|black|hispanic"))) %>% 
  map2(mod_names, ~ mutate(.x, model = .y)) %>% 
  map(~ mutate(.x, term = case_when(
    term == "threat_overall_scaled" ~ "Proximal Exposure",
    term == "black_share_btw_traditional_scaled" ~ "Direct (%Black - TPS)",
    term == "hispanic_share_btw_traditional_scaled" ~ "Direct (%Hispanic - TPS)"
  )))

tidy_models %>% 
  bind_rows() %>% 
  dwplot() +
  geom_vline(
    xintercept = 0, 
    colour = "grey60", 
    linetype = 2) + 
  scale_color_brewer(name = "", palette = "Paired", direction = -1) + 
  labs(y = "", x = "Coefficient Estimate",
       title = "Comparing white charter enrollment in non-urban and urban districts",
       caption = str_wrap("TPS = Traditional public schools. Coefficient estimates based on OLS regression with district and year fixed effects. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of children in poverty; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, white students, and magnet schools in the district; and teacher-student ratio.", 80)) +
  coord_flip()

tab <- huxreg(models, 
       coefs = c("Proximal threat" = "threat_overall_scaled", 
                 "%Black(TPS)" = "black_share_btw_traditional_scaled",
                 "%Hispanic(TPS)" = "hispanic_share_btw_traditional_scaled"),
       statistics = c('N' = 'nobs', 'R-squared' = 'r.squared'),
       note = "{stars}. TPS = Traditional public schools. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of 5-17 in poverty; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, and magnet schools in the district; and teacher-student ratio.") %>% 
  insert_row(c("District Fixed Effects", "X", "X", "X", "X"), after = 7) %>% 
  insert_row(c("Year Fixed Effects", "X", "X", "X", "X"), after = 8) %>% 
  insert_row(c("Controls", "X", "X", "X", "X"), after = 9) %>% 
  set_bottom_border(8, 2:5, 0) %>% 
  set_bottom_border(9, 2:5, 0) %>% 
  set_bottom_border(10, 2:5, 0) %>% 
  insert_row(c("", "Share of Charters", "", "White Enrollment in Charters", ""), after = 0) %>% 
  set_colspan(1, c(2, 4), 2) %>% 
  set_bottom_border(1, 1:5, 1) %>% 
  set_bottom_border(1, 1:5, 0) %>% 
  set_top_border(2, 1:5, 0) %>%
  set_align(1:14, 2:5, "center") %>% 
  insert_row(c("Table 2: Results for urban and non-urban school districts", 
               "", "", "", ""), after = 0) %>% 
  set_colspan(1, 1, 5) %>% 
  set_top_border(1, 1:5, 0) %>% 
  set_align(1, 1, "left") %>% 
  set_italic(1, 1, TRUE) %>% 
  set_wrap(1, 1, TRUE)
```

```{r out.width='90%', fig.align='center', dpi=340}
m1 <- 
  felm(
    isolation_scaled ~
      black_overall_district_share + 
      hispanic_overall_district_share + 
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct | leaid + year |  (white_share_within_charter_scaled ~ threat_overall_scaled) | leaid,
    data = charter_dat
  )

m2 <- 
  felm(
    isolation_scaled ~
      black_overall_district_share + 
      hispanic_overall_district_share + 
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct | leaid + year |  (white_share_within_charter_scaled ~  hispanic_share_btw_traditional_scaled) | leaid,
    data = charter_dat
  )

m3 <- 
  felm(
    isolation_scaled ~
      black_overall_district_share + 
      hispanic_overall_district_share + 
      t_s_ratio + 
      share_swd_scaled + 
      share_title_i_scaled + 
      share_magnet_scaled + 
      log(enrollment) +
      log(rev_local_total) +
      log(per_pupil_spending) +
      est_population_5_17_poverty_pct | leaid + year |  (white_share_within_charter_scaled ~ threat_overall_scaled + black_share_btw_traditional_scaled + hispanic_share_btw_traditional_scaled) | leaid,
    data = charter_dat
  )

models <- list("Proximal Exposure" = m1, "Direct Exposure" = m2, 
               "Proximal and Direct Exposure" = m3)
mod_names <- names(models)
tidy_models <- 
  map(models, tidy) %>% 
  map(~ filter(.x, str_detect(term, "fit"))) %>% 
  map2(mod_names, ~ mutate(.x, model = .y)) %>% 
  map(~ mutate(.x, term = "Fitted Share of White Students in Charters"
  ))

tidy_models %>% 
  bind_rows() %>% 
  dwplot() +
  geom_vline(
    xintercept = 0, 
    colour = "grey60", 
    linetype = 2) + 
  scale_color_brewer(name = "", palette = "Paired", direction = -1) + 
  labs(y = "", x = "Coefficient Estimate",
       title = "Racial Isolation and Threat-induced White Enrollment in Charters",
       caption = str_wrap("Coefficient estimates based on IV regression with district and year fixed effects. Standard errors clustered at the district level. Controls include the natural log of enrollment, local revenue, and per-student spending; the share of children in poverty; the share of Title 1 eligible schools, students with disabilities, and magnet schools in the district; and teacher-student ratio.", 80)) +
  coord_flip()
```




# Conclusion

Try `posterdown` out! Hopefully you like it!

```{r, include=FALSE}
knitr::write_bib(c('knitr','rmarkdown','posterdown','pagedown'), 'packages.bib')
```

# References
